---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# rfishprod

<!-- badges: start -->
<!-- badges: end -->

Implements a framework to generate individual-level estimates
    of fish productivity, with a focus on reef fishes. This individual
    approach works by combining estimates of somatic growth through the
    von Bertalanffy Growth Model, and deterministic or stochastic natural
    mortality using instantaneous mortality rates.

## Installation

You can install rfishprod from [GitHub](https://github.com/renatoamorais/rfishprod) with:

``` r
devtools::install_github("renatoamorais/rfishprod")
```

To get [`devtools`](https://cran.r-project.org/web/packages/devtools/index.html), simply use `install.packages("devtools")`.

Although not a dependency, I recommend using R 3.6.3 or higher. This is mainly because [`xgboost`]( https://CRAN.R-project.org/package=xgboost), the machine behind internal prediction, demands R 3.6.3 in its most recent version.

## NEWS!

Recently (August 2022), I started working on the package again. Mostly updated documentation and changed a couple of functions (see below) with more sound approaches, but nothing that changes for the user. Might do more updates soon, including a vignette.

## Recent updates

On ***08.08.2022***: included an alternative parameterisation of the VBGM, the original one specifying the intercept (L0 or length at age 0). This is a biologically intuitive and more relevant form. It also ends arbitrarities of the rare instances when surveyed fishes were smaller than the estimated L0 from the model. This more natural form is recommended, but requires specifying for each fish its settlement size. In practice it is just another column in the data. If L0 is not specified, model keeps running business as usual.

On ***02.08.2022***: included the empirical equation to estimate instant mortality from Lorenzen et al (2022) Fish Res 252:106327. Also changed the method = 'Function' to leave the exponent of the relationship between mortality and body length adjustable by the user (it was fixed as -0.75) given evidence of the same paper that this exp could be closer to -1 (value of -0.91 set as default, see Table 2, model 3 in this ref).


## Example


```{r example, eval = F}
library(rfishprod)

# Check dataset repdata #
(repdata <- rfishprod:::repdata)

# Getting levels ready #
repdata <- tidytrait(repdata, db)

# Formula from Morais and Bellwood (2018) #
fmod <- formula(~ sstmean + MaxSizeTL + Diet + Position + Method)

# Predicting Kmax, the standardised VBGF parameter (Recommendation: use 100s to 1000s iterations) #
datagr <- predKmax(repdata, 
                   dataset = db,
                   fmod = fmod,
                   niter = 10,
                   return = 'pred')

datagr <- datagr$pred

# Predicting M/Z: the instantaneous mortality rate (Recommendation: see help file for) #
# Using the Lorenzen equation
datagr$Md <- with (datagr, 
					predM (Lmeas = Size, 
						     Lmax = MaxSizeTL, 
						     Kmax = Kmax, 
						     method = 'Lorenzen'))
# But also the Gislason one
datagr$Md <- with (datagr, 
					predM (Lmeas = Size, 
						     Lmax = MaxSizeTL, 
						     Kmax = Kmax, 
						     method = 'Gislason'))
						   

# Positioning your fish in their growth trajectory #
# aka. what's the size they're supposed to have on the next day? #
# now using L0, i.e., size at settlement, to constrain the curve #
with(datagr, applyVBGF(Lmeas = Size,
                       Lmax = MaxSizeTL,
                       Kmax = Kmax,
                       L0   = L0))
						 
# Compare with their size on the previous day #
datagr$Size

# Estimating gross somatic growth (g) #
with(datagr, somaGain(a = a,
                      b = b,
                      Lmeas = Size,
                      Lmax = MaxSizeTL,
                      Kmax = Kmax))
						      
# Applying stochastic mortality #
applyMstoch(datagr$Md)


# Alternatively, estimating per capita mass loss due to mortality #
with(datagr, somaLoss(M = Md,
                      Lmeas = Size,
                      a = a,
                      b = b))

```

## Citation

Please, if you're using `rfishprod`, the relevant citation for the package can be obtained from:

```{r, eval = FALSE}
citation("rfishprod")
```


## Issues

Please [report issues or bugs](https://github.com/renatoamorais/rfishprod/issues) or shoot me an email (just hit `?rfishprod`)